---
title: "Calculating Vertebrate Metabolic Rates from Blyth et al. (2026)"
author: ""
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Overview

Blyth et al. (2026, *Ecology Letters*) updated allometric normalisation constants for
metabolism and production using >1000 genera and Bayesian Phylogenetic Least Squares
(PGLS). This document explains how to use their Table 1 intercept estimates to predict
**basal** and **field** metabolic rates (whole-body or mass-specific) for an unknown
vertebrate, classified as either ectotherm or endotherm.

---

## The General Allometric Equation

Metabolic rate scales with body mass and temperature as (Blyth et al. Eq. 1):

$$X \, [\text{W}] = e^{C} \cdot M^{b} \cdot e^{-E_a / kT}$$

For the **mass-specific** rate (Blyth et al. Eq. 2):

$$x \, [\text{W g}^{-1}] = e^{C} \cdot M^{b-1} \cdot e^{-E_a / kT}$$

This is derived from the PGLS regression fitted on temperature-corrected data:

$$\ln\!\left(X \cdot e^{E_a / kT}\right) = C + b \cdot \ln(M)$$

### Parameters

| Symbol | Description | Value / Source |
|--------|-------------|----------------|
| $C$ | Allometric intercept in ln(W) | From Table 1 (Blyth column) |
| $M$ | Body mass | **grams** |
| $b$ | Allometric scaling exponent | From Table S3 (varies by group; ~0.75 if unavailable) |
| $E_a$ | Activation energy | 0.63 eV |
| $k$ | Boltzmann's constant | 8.617 × 10⁻⁵ eV K⁻¹ |
| $T$ | Temperature | Kelvin — body temp (endotherms), environmental temp (ectotherms) |

**Temperature defaults for endotherms** (if body temperature is unknown):
mammals = 37°C (310 K); birds = 39°C (312 K).

---

## Table 1 Intercepts for Vertebrates

The PGLS intercepts ($C$, in ln(W)) from Blyth et al. Table 1 for the two vertebrate
groups relevant to a mystery vertebrate:

```{r intercepts}
intercepts <- data.frame(
  group            = c("Ectothermic vertebrate", "Endothermic vertebrate"),
  C_blyth          = c(17.4,  19.53),
  CRI_lower        = c(16.46, 18.93),
  CRI_upper        = c(18.28, 20.13),
  C_brose          = c(18.18, 19.5),   # Previous standard (Brose et al. 2006)
  ax_brose         = c(0.88,  3.22),   # Brose BEFW coefficient
  ax_blyth         = c(0.274, 2.27)    # Blyth BEFW coefficient
)

knitr::kable(
  intercepts[, c("group", "C_blyth", "CRI_lower", "CRI_upper", "C_brose",
                 "ax_brose", "ax_blyth")],
  col.names = c("Group", "C Blyth [ln(W)]", "95% CRI lower",
                "95% CRI upper", "C Brose [ln(W)]",
                "a_x Brose", "a_x Blyth"),
  digits = 3,
  caption = "Table 1 intercepts and BEFW coefficients from Blyth et al. (2026).
             Blyth intercepts are lower than Brose for ectothermic vertebrates,
             reflecting phylogenetic correction and broader sampling."
)
```

Note that the Blyth intercepts are **lower** than the Brose et al. (2006) values for
ectothermic vertebrates (17.4 vs. 18.18 ln(W)), reflecting the phylogenetic correction
and broader taxonomic sampling. For endotherms the estimates are similar.

---

## Basal vs. Field Metabolic Rate

The data used by Blyth et al. are:

- **BMR** (basal metabolic rate) for endotherms
- **SMR** (standard metabolic rate) for ectotherms

Both are **resting/basal** rates. To obtain **field metabolic rate (FMR)**, multiply by
3, following Nagy (1987) as cited in Blyth et al.:

$$\text{FMR} = 3 \times \text{BMR (or SMR)}$$

This ×3 factor is already incorporated into the BEFW model coefficients ($a_x$) in
Table 1, but is **not** in the raw intercept $C$. When using $C$ directly in the
equation above, you obtain the resting rate; multiply by 3 for FMR.

---

## R Implementation

### Core functions

```{r functions}
k_bolt <- 8.617e-5  # Boltzmann constant (eV/K)
Ea     <- 0.63      # Activation energy (eV); Brown et al. 2004

#' Whole-body metabolic rate in Watts (BMR for endotherms, SMR for ectotherms)
#'
#' @param M    Body mass in grams
#' @param C    Allometric intercept in ln(W) from Table 1
#' @param b    Allometric exponent (use ~0.75 if Table S3 unavailable)
#' @param T_K  Temperature in Kelvin
#' @return Metabolic rate in Watts
bmr_smr_W <- function(M, C, b, T_K) {
  exp(C) * M^b * exp(-Ea / (k_bolt * T_K))
}

#' Mass-specific metabolic rate in W/g
#'
#' @inheritParams bmr_smr_W
#' @return Mass-specific metabolic rate in W/g
bmr_smr_mass_specific <- function(M, C, b, T_K) {
  exp(C) * M^(b - 1) * exp(-Ea / (k_bolt * T_K))
}

#' Field metabolic rate (whole-body, Watts)
fmr_W <- function(M, C, b, T_K) {
  3 * bmr_smr_W(M, C, b, T_K)
}

#' Field metabolic rate (mass-specific, W/g)
fmr_mass_specific <- function(M, C, b, T_K) {
  3 * bmr_smr_mass_specific(M, C, b, T_K)
}
```

### Group-specific parameters

```{r params}
# Intercepts from Table 1 (Blyth et al. 2026, PGLS median)
C_ecto_vert <- 17.4
C_endo_vert <- 19.53

# Exponent b: from Table S3 (supplementary). If unavailable, use classical ~0.75.
# Replace these with Table S3 values when available.
b_ecto_vert <- 0.75
b_endo_vert <- 0.75

# Reference temperatures (Kelvin)
T_mammal <- 273.15 + 37  # 310.15 K
T_bird   <- 273.15 + 39  # 312.15 K
T_ecto   <- 273.15 + 20  # 293.15 K — example environmental temperature
```

### Example predictions

```{r examples}
library(dplyr)
library(tidyr)
library(ggplot2)

# Define a range of body masses (grams)
masses_g <- 10^seq(-2, 6, by = 0.25)  # 0.01 g to 1,000 kg

results <- bind_rows(
  # Ectothermic vertebrate at 20°C
  data.frame(
    M_g        = masses_g,
    group      = "Ectotherm (20°C)",
    BMR_W      = bmr_smr_W(masses_g, C_ecto_vert, b_ecto_vert, T_ecto),
    FMR_W      = fmr_W(masses_g, C_ecto_vert, b_ecto_vert, T_ecto)
  ),
  # Endothermic vertebrate — mammal (37°C)
  data.frame(
    M_g        = masses_g,
    group      = "Endotherm — mammal (37°C)",
    BMR_W      = bmr_smr_W(masses_g, C_endo_vert, b_endo_vert, T_mammal),
    FMR_W      = fmr_W(masses_g, C_endo_vert, b_endo_vert, T_mammal)
  ),
  # Endothermic vertebrate — bird (39°C)
  data.frame(
    M_g        = masses_g,
    group      = "Endotherm — bird (39°C)",
    BMR_W      = bmr_smr_W(masses_g, C_endo_vert, b_endo_vert, T_bird),
    FMR_W      = fmr_W(masses_g, C_endo_vert, b_endo_vert, T_bird)
  )
)

# Print table for a few representative masses
representative <- results %>%
  filter(round(log10(M_g), 2) %in% c(-2, 0, 2, 4, 6)) %>%
  mutate(
    M_kg    = M_g / 1000,
    BMR_mW  = BMR_W * 1000,
    FMR_mW  = FMR_W * 1000
  ) %>%
  select(group, M_kg, BMR_mW, FMR_mW)

knitr::kable(
  representative,
  col.names = c("Group", "Mass (kg)", "BMR/SMR (mW)", "FMR (mW)"),
  digits    = 3,
  caption   = "Predicted whole-body basal/standard (BMR/SMR) and field (FMR)
               metabolic rates at representative body masses using Blyth et al.
               (2026) intercepts and b = 0.75."
)
```

### Plot: metabolic rate vs. body mass

```{r plot, fig.width=8, fig.height=5}
results_long <- results %>%
  pivot_longer(cols = c(BMR_W, FMR_W), names_to = "rate_type", values_to = "rate_W") %>%
  mutate(
    rate_type = recode(rate_type, BMR_W = "BMR/SMR", FMR_W = "FMR"),
    M_kg = M_g / 1000
  )

ggplot(results_long, aes(x = M_kg, y = rate_W, colour = group, linetype = rate_type)) +
  geom_line(size = 0.9) +
  scale_x_log10(
    labels = scales::trans_format("log10", scales::math_format(10^.x))
  ) +
  scale_y_log10(
    labels = scales::trans_format("log10", scales::math_format(10^.x))
  ) +
  scale_linetype_manual(values = c("BMR/SMR" = "solid", "FMR" = "dashed")) +
  labs(
    x        = "Body mass (kg)",
    y        = "Metabolic rate (W)",
    colour   = "Group",
    linetype = "Rate type",
    title    = "Predicted vertebrate metabolic rates",
    subtitle = "Blyth et al. (2026) intercepts; b = 0.75; FMR = 3 × BMR/SMR"
  ) +
  theme_bw(base_size = 13)
```

---

## Propagating Uncertainty from the CRI

The 95% credible intervals on $C$ from the Bayesian PGLS can be directly used to
bracket uncertainty on predictions:

```{r uncertainty, fig.width=8, fig.height=5}
# Ectothermic vertebrate example with CRI propagation
C_ecto_low  <- 16.46
C_ecto_high <- 18.28

df_uncertainty <- data.frame(M_g = masses_g) %>%
  mutate(
    BMR_median = bmr_smr_W(M_g, C_ecto_vert, b_ecto_vert, T_ecto),
    BMR_low    = bmr_smr_W(M_g, C_ecto_low,  b_ecto_vert, T_ecto),
    BMR_high   = bmr_smr_W(M_g, C_ecto_high, b_ecto_vert, T_ecto),
    M_kg       = M_g / 1000
  )

ggplot(df_uncertainty, aes(x = M_kg)) +
  geom_ribbon(aes(ymin = BMR_low, ymax = BMR_high), alpha = 0.25, fill = "orange") +
  geom_line(aes(y = BMR_median), colour = "orange", size = 1) +
  scale_x_log10(labels = scales::trans_format("log10", scales::math_format(10^.x))) +
  scale_y_log10(labels = scales::trans_format("log10", scales::math_format(10^.x))) +
  labs(
    x        = "Body mass (kg)",
    y        = "SMR (W)",
    title    = "Ectothermic vertebrate SMR with 95% CRI uncertainty",
    subtitle = "Shaded band spans the 95% credible interval on intercept C (Table 1)"
  ) +
  theme_bw(base_size = 13)
```

---

## Step-by-Step Recipe

For a mystery vertebrate of known body mass and approximate temperature:

1. **Classify** as ectotherm or endotherm
2. **Choose $C$** from Table 1 (Blyth median; use CRI for uncertainty bounds):
   - Ectothermic vertebrate: $C = 17.4$ [95% CRI: 16.46–18.28]
   - Endothermic vertebrate: $C = 19.53$ [95% CRI: 18.93–20.13]
3. **Set $b$** from Table S3 of Blyth et al. (2026); use $b \approx 0.75$ if unavailable
4. **Set $T$** in Kelvin:
   - Mammals: 310.15 K (37°C); Birds: 312.15 K (39°C)
   - Ectotherms: environmental temperature
5. **Compute resting rate** (W): $X = e^C \cdot M^b \cdot e^{-E_a/kT}$
6. **For FMR**: multiply by 3
7. **For mass-specific**: use exponent $b - 1$ in place of $b$

---

## Important Caveats

- The **exponent $b$** is in Table S3 (supplementary only); $b \approx 0.75$ is a
  reasonable fallback but the paper did not fix this — group-specific values will
  differ.
- The Blyth intercepts are generally **lower** than Brose et al. (2006), predicting
  lower metabolic rates and consequently higher biomass and species persistence in
  food web models.
- **Endothermy is not always binary**: some fish (e.g., opah, bluefin tuna) exhibit
  regional endothermy and may not fit cleanly into the ectothermic vertebrate category.
- The **BEFW model coefficients** $a_x$ in Table 1 (0.274 for ecto vert, 2.27 for
  endo vert) are *not* the same as $e^C$ — they have been further transformed via a
  unit conversion factor ($C_\text{fac} = 51.7$) and the FMR ×3 factor for use in
  the bioenergetic food web model. Use the raw intercepts $C$ for predicting
  metabolic rates in Watts.

---

## References

- Blyth, P.S.A. et al. (2026). The Critical Role of Coefficients: Updating Allometric
  Normalisation Constants for Modern Ecology and Modelling. *Ecology Letters*, 29,
  e70330. https://doi.org/10.1111/ele.70330
- Brose, U. et al. (2006). Consumer–resource body-size relationships in natural food
  webs. *Ecology*, 87, 2411–2417.
- Nagy, K.A. (1987). Field metabolic rate and food requirement scaling in mammals and
  birds. *Ecological Monographs*, 57, 111–128.
- Brown, J.H. et al. (2004). Toward a metabolic theory of ecology. *Ecology*, 85,
  1771–1789.
